Confirmation of \srcfull{} as a long period radio transient

### 2025-05-26 paper/main.tex
Latest synch of paper with Overleaf version -- including more instances of addressing co-authors' comments

% Abstract of the paper
\begin{abstract}
We report the discovery of several new pulses from the source \srcfull{} (\src{}), confirming it as a long period transient (LPT) with a period of ${\sim}1.16\,$hours. The pulses are significantly scattered, consistent with Galactic electron density models. Two of the new pulses also had measurable polarisation, but unlike the originally detected pulse, the polarisation angle does not behave as expected from the rotating vector model. We interpret historical non-detections of \src{} as an intrinsic intermittency that occurs on month-long timescales, and discuss possible causes. We conjecture that, like other LPTs with periods $\gtrsim 1\,$hour, \src{} is a white-dwarf binary pulsar, but note that its period is marginally shorter than the minimum known orbital period for the class of cataclysmic variables known as polars.
\end{abstract}

@@

\srcfull{} (hereafter, \src{}) is a transient radio source discovered in the Variables and Slow Transients (VAST) survey \citep[][hereafter \citetalias{2024MNRAS.535..909D}]{2024MNRAS.535..909D}.
A single, highly polarised pulse was observed, lasting approximately 2 minutes, exhibiting a steep spectrum ($\alpha = -3.1$, for flux density $S_\nu\propto\nu^\alpha$).
Extensive follow up in new and archival data at multiple wavelengths turned up no other pulses.
Despite this, the authors concluded that \src{} was most likely a long period transient (LPT), a newly emerging class of radio source with periods ranging from minutes to hours \citep[see e.g.][]{2022Natur.601..526H,2023Natur.619..487H,2024NatAs...8.1159C}.

Although optical follow-up has confirmed that at least some LPTs are a type of strongly magnetic cataclysmic variable (CV) known as polars \citep{deRuiter2025,2024ApJ...976L..21H,2025A&A...695L...8R}, it remains unclear if this model can account for all known LPTs \citep[e.g.][]{2022ApJ...940...72R,Lee2025}.
Optical follow-up is challenging for \src{}, which is situated very close to the Galactic plane ($b = -0\overset{\circ}{.}12$), at a distance of ${\sim}4.7\,$kpc \citepalias{2024MNRAS.535..909D}.
Unsurprisingly, archival searches at \src{}'s location did not turn up any source down to ${\sim}23$ AB mag.
Without more detections in the radio or at other wavelengths, it seemed unlikely that the nature of \src{} could ever be resolved.

@@

The MWA, ASKAP and MeerKAT are all radio interferometers capable of imaging.
A summary of all the observations included in this work is summarised in \Tab~\ref{tbl:obs}.
In this section, we describe how each telescope was used to generate dynamic spectra of \src{}.

\begin{table*}
  \centering
  \caption{Observations summary}
  \label{tbl:obs}
  \begin{tabular}{lllccc}
    \hline
    Date range & Telescope & Project & $\nu$ & $\Delta\nu$ & Number of \\ 
    (MJD) & & & (MHz) & (MHz) & pulses \\
    \hline 
    59965 & ASKAP & SB47253 (original discovery) & 887.5 & 288 & 1 \\
    59966 & MWA & G0060 & 161.92 & 15.36 & 1 \\
    %60084 & MWA & G0057 & 154.24 & 30.72 & $\le$ 1 \\
    60040  & ASKAP & SB49153 & 887.5 & 288 & 1 \\
    60092 - 60093 & MeerKAT & DDT-20230525-DD-01 & 1284 & 856 & 4 \\
    60463 - 60563 & MWA & G0080 & 200.32 & 30.72 & 16 \\
    60503 & ASKAP & SB63600 & 887.5 & 288 & 1 \\
    60572 & MWA & D0042 & 184.96 & 30.72 & 4 \\
    60592 - 60602 & MWA & D0042 & 200.32 & 30.72 & 17 \\
    60602 & MeerKAT & DDT-20241015-NH-01 & 812.8 & 495.3 & 1 \\
    \hline
  \end{tabular}
\end{table*}

@@

\citetalias{2024MNRAS.535..909D} presented an exhaustive search of the ASKAP data available at the time of writing, comprising 60\,h of data, and yielding only a single detection of \src{} (in Scheduling Block ID 47253). Subsequent to the writing of \citetalias{2024MNRAS.535..909D}, a further epoch of the VAST survey was performed (SBID 63600) in which \src{} produced a detectable pulsation.

@@

No significant polarised emission was detected in the MWA pulses, as well as the 2023 MeerKAT pulses.
The apparent lack of polarisation in the 2023 MeerKAT data set is likely due to low S/N coupled with the source not being centred in the beam.
Much more careful imaging analysis would be needed to obtain reliable measurements or limits on the polarised components of those pulses.

The apparent lack of linear polarisation in the MWA pulses, however, is expected due to the high rotation measure (RM) of ${\sim}961\,{\rm rad}/{\rm m}^2$ \citepalias[as reported in][]{2024MNRAS.535..909D}, causing the pulses to be depolarised within individual channels.
We also report that no significantly circularly polarised emission was seen at MWA frequencies.

The three sufficiently bright pulses for which we report significant polarisation are (1) the original 2023 ASKAP pulse \citepalias{2024MNRAS.535..909D}, (2) the 2024 ASKAP pulse, and (3) the 2024 MeerKAT pulse.
We applied the parallactic angle correction to the MeerKAT data, as this had not already been applied during upstream processing.

@@

\src{}'s period and duty cycle are not dissimilar to ILT J1101+5521 and GLEAM-X J0704$-$37, which are confirmed polars (\Fig~\ref{fig:lpt_comparison}), suggesting the possibility that \src{}, too, is a polar.
However, its period of ${\sim}1.16\,$hours would make it shorter than the minimum known orbital period of $1.3\,$hours for confirmed polars \citep{schwope2025polarcatcatalogpolarslowaccretion}.
This makes it an interesting test bed for the suggestion advanced by \citet{2025A&A...695L...8R} that `short LPTs' ($P \lesssim 78\,$mins) and `long LPTs' ($P \gtrsim 78\,$mins) represent different classes of systems (AM Canum Venaticorum stars and CVs, respectively).
\begin{figure*}
    \centering
    \includegraphics[width=0.98\linewidth]{lpt_comparison.pdf}
    \caption{Duty cycle vs radio pulse period for various long period radio emitters \citep{deRuiter2025,2022Natur.601..526H,2023Natur.619..487H,2024NatAs...8.1159C,2005Natur.434...50H,deRuiter2025,Lee2025,2024arXiv241116606W}. The duty cycles are derived from the reported pulse widths for LPTs, which are compared to the range of the fitted, unscattered pulse widths of \src{} at 10\% of the peak. AR\,Sco \citep{2016Natur.537..374M,2018A&A...611A..66S} and other sources without well-defined duty cycles or periods \citep[e.g.][and references therein]{2021ApJ...920...45W} are not included on this plot.}
    \label{fig:lpt_comparison}
\end{figure*}

@@

Although there is as yet no clear consensus on whether LPTs represent one class or multiple classes of system \citep[e.g.][]{2024ApJ...961..214R}, the recent confirmation of J1912$-$4410 as a pre-evolved white dwarf pulsar \citep{2023NatAs...7..931P} and ILT\,J1101+5521 and GLEAM-X\,J0704$-$37 as polars \citep{deRuiter2025,2025A&A...695L...8R} raise the intriguing possibility that other LPTs may also be radio-emitting white dwarf pulsars at various stages of evolution (\citealt{2021NatAs...5..648S}, Horv\'{a}th et al. in prep).

>>>>

% Abstract of the paper
\begin{abstract}
We report the discovery of several new pulses from the source \srcfull{} (\src{}), originally identified from a single 2-min log pulse, confirming it as a long period transient (LPT) with a period of ${\sim}1.16\,$hours. The pulses are significantly scattered, consistent with Galactic electron density models. Two of the new pulses also had measurable polarisation, but unlike the originally detected pulse, the polarisation angle does not behave as expected from the rotating vector model. We interpret historical non-detections of \src{} as an intrinsic intermittency that occurs on month-long timescales, and discuss possible causes. We conjecture that, like other LPTs with periods $\gtrsim 1\,$hour, \src{} is a white-dwarf binary pulsar, but note that its period is marginally shorter than the minimum known orbital period for the class of cataclysmic variables known as polars.
\end{abstract}

@@

\srcfull{} (hereafter, \src{}) is a transient radio source discovered \citep[][hereafter \citetalias{2024MNRAS.535..909D}]{2024MNRAS.535..909D} in the Variables and Slow Transients (VAST) survey \citep{2013PASA...30....6M}.
A single, highly polarised pulse was observed, lasting approximately 2 minutes, exhibiting a steep spectrum ($\alpha = -3.1$, for flux density $S_\nu\propto\nu^\alpha$).
Extensive follow up in new and archival data at multiple wavelengths turned up no other pulses.
Despite this, the authors concluded that \src{} was most likely a long period transient (LPT), a newly emerging class of radio source with periods ranging from minutes to hours \citep[see e.g.][]{2022Natur.601..526H,2023Natur.619..487H,2024NatAs...8.1159C}.

Although optical observations have provided compelling evidence that at least some LPTs are detached, possibly strongly magnetic, white dwarf (WD) binaries \citep{deRuiter2025,2024ApJ...976L..21H,2025A&A...695L...8R}, it remains unclear if this model can account for all known LPTs \citep[e.g.][]{2022ApJ...940...72R,Lee2025}.
Optical follow-up is challenging for \src{}, which is situated very close to the Galactic plane ($b = -0\overset{\circ}{.}12$), at a distance of ${\sim}4.7\,$kpc \citepalias{2024MNRAS.535..909D}.
Unsurprisingly, archival searches at \src{}'s location did not turn up any source down to ${\sim}23$ AB mag.
Without more detections in the radio or at other wavelengths, it seemed unlikely that the nature of \src{} could ever be resolved.

@@

Pulses from \src{} were detected with the Murchison Widefield Array \citep[MWA;][]{Tingay2013}, the Australian SKA Pathfinder \citep[ASKAP;][]{2021PASA...38....9H} and MeerKAT \citep{2016mks..confE...1J}, are all radio interferometers capable of imaging.
A summary of all the observations included in this work is summarised in \Tab~\ref{tbl:obs}.
In this section, we describe how each telescope was used to generate dynamic spectra of \src{}.

\begin{table*}
  \centering
  \caption{Observations summary}
  \label{tbl:obs}
  \begin{tabular}{lllcccc}
    \hline
    Date range & Telescope & Project & $\nu$ & $\Delta\nu$ & Obs. length & Number of \\ 
    (MJD) & & & (MHz) & (MHz) & (min) & pulses \\
    \hline 
    59965 & ASKAP & SB47253 (original discovery) & 887.5 & 288 & 12.1 & 1 \\
    59966 & MWA & G0060 & 161.92 & 15.36 & 6.7 & 1 \\
    %60084 & MWA & G0057 & 154.24 & 30.72 & N/A & $\le$ 1 \\
    60040  & ASKAP & SB49153 & 887.5 & 288 & 12.1 & 1 \\
    60092 - 60093 & MeerKAT & DDT-20230525-DD-01 & 1284 & 856 & 307 & 4 \\
    60463 - 60563 & MWA & G0080 & 200.32 & 30.72 & (Each obs.) 4.9 & 16 \\
    60503 & ASKAP & SB63600 & 887.5 & 288 & 12.3 & 1 \\
    60572 & MWA & D0042 & 184.96 & 30.72 & (Each obs.) 4.9 & 4 \\
    60592 - 60602 & MWA & D0042 & 200.32 & 30.72 & (Each obs.) 4.9 & 17 \\
    60602 & MeerKAT & DDT-20241015-NH-01 & 812.8 & 495.3 & 10 & 1 \\
    \hline
  \end{tabular}
\end{table*}

@@

\citetalias{2024MNRAS.535..909D} presented an exhaustive search of the ASKAP data available at the time of writing, comprising 60\,h of data, and yielding only a single detection of \src{} (in Scheduling Block ID 47253). Subsequent to the writing of \citetalias{2024MNRAS.535..909D}, a further search for transient signals in ASKAP data was performed (SBID 63600) in which \src{} produced a detectable pulsation.

@@

No significant linear or circular polarised emission was detected in the 2023 MeerKAT pulses.
This is likely due to low S/N coupled with the source not being centred in the beam.
Much more careful imaging analysis would be needed to obtain reliable measurements or limits on the polarised components of those pulses.

Given \src{}'s rotation measure (RM) of ${\sim}961\,{\rm rad}/{\rm m}^2$ \citepalias[as reported in][]{2024MNRAS.535..909D}, linearly polarised emission would undergo several tens of degrees of rotation per 40\,kHz channel of the GPM observations, and therefore be somewhat depolarised.
To offer the best chance of detecting polarised emission, the MWA DDT observations between MJD 60592 and 60602 (see \Tab~\ref{tbl:obs}) were taken with a frequency resolution of 10\,kHz, which would result in only $12^\circ$ or rotation per channel at 200\,MHz.
We report no significant linearly polarised emission after de-Faraday rotating\footnote{On account of the possible sign ambiguity of the reported RM, we also tried de-Faraday rotating the MWA observations at $961\,{\rm rad}/{\rm m}^2$.} these observations at $-961\,{\rm rad}/{\rm m}^2$.
We also report that no significantly circularly polarised emission was seen in any MWA observation.

The three sufficiently bright pulses for which we report significant polarisation are (1) the original 2023 ASKAP pulse, whose polarisation has already been discussed at length in \citetalias{2024MNRAS.535..909D}, (2) the 2024 ASKAP pulse, and (3) the 2024 MeerKAT pulse.
We applied the parallactic angle correction to the MeerKAT data, as this had not already been applied during upstream processing.

@@

\src{}'s period and duty cycle are not dissimilar to ILT J1101+5521 and GLEAM-X J0704$-$37, which are conjectured to be on the evolutionary path towards becoming polars (\Fig~\ref{fig:lpt_comparison}), suggesting the possibility that \src{}, too, is such a system.
Notably, its period of ${\sim}1.16\,$hours would make it shorter than the minimum known orbital period of $1.3\,$hours for confirmed polars \citep{schwope2025polarcatcatalogpolarslowaccretion}.
This makes it an interesting test bed for the possibility that LPTs with different periods represent different stages of WD binary evolution, as they progress towards becoming polars.
\begin{figure*}
    \centering
    \includegraphics[width=0.98\linewidth]{lpt_comparison.pdf}
    \caption{Duty cycle vs radio pulse period for various long period radio emitters \citep{deRuiter2025,2022Natur.601..526H,2023Natur.619..487H,2024NatAs...8.1159C,2005Natur.434...50H,deRuiter2025,Lee2025,2024arXiv241116606W}. The duty cycles are derived from the reported pulse widths for LPTs, which are compared to the range of the fitted, unscattered pulse widths of \src{} at 10\% of the peak. AR\,Sco \citep{2016Natur.537..374M,2018A&A...611A..66S} and other sources without well-defined duty cycles or periods \citep[e.g.][and references therein]{2021ApJ...920...45W} are not included on this plot.}
    \label{fig:lpt_comparison}
\end{figure*}

@@

Although there is as yet no clear consensus on whether LPTs represent one class or multiple classes of system \citep[e.g.][]{2024ApJ...961..214R}, the recent confirmation of J1912$-$4410 as a pre-evolved white dwarf pulsar \citep{2023NatAs...7..931P} and ILT\,J1101+5521 and GLEAM-X\,J0704$-$37 as likely candidates for polars or polar-like systems \citep{deRuiter2025,2025A&A...695L...8R} raise the intriguing possibility that other LPTs may also be radio-emitting white dwarf pulsars at various stages of evolution (\citealt{2021NatAs...5..648S}, Horv\'{a}th et al. in prep).
